---
title: "README"
author: "Sophia Hess"
date: "2025-04-28"
output: md_document
github_document:
pandoc_args: ["--wrap=none"]
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
install.packages("corrplot", repos = "http://cran.us.r-project.org")

# Load necessary libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(corrplot)
library(gridExtra)
```

## -- Business Understanding --
## What makes a good quality wine? 🍷
Based on the research, wine producers want to identify controllable factors that could 
help improve product quality, particularly in chemical composition of wine. 
A high-quality wine is influenced by a combination of chemical, sensory, and structural 
attributes. Research and industry analysis highlights factors like **alcohol content**, 
**acidity**, **sugar levels**, and **volatile acidity** as crucial to consumer perception and 
expert ratings. Higher alcohol levels are linked to better body and richness, and 
lower volatile acidity is linked to cleaner, more pleasant aromas. **Citric acid** also 
contributes to a wine's freshness, and **sulphates** help preserve wine and enhance structure. 
Additionally, **pH levels** and **density** affect the wine's stability, fermentation quality, and 
taste balance. Understanding how all these variables interact allows producers to refine 
fermentation and aging processes to make more desirable wines.

## -- Data Understanding --
## General info about the dataset
1. **Dataset:** UCI Wine Quality Dataset
2. **Samples:** 1,599 red wines
3. **Features:** 11 numeric chemical properties + 1 quality rating
4. **Target variable:** `quality` (integer from 0-10, treated as ordinal)
5. **Feature types:** all features are **ratio** (ex/ pH, alcohol, sulphates)

```{r load_data}
# Download red wine dataset
wine <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv", sep = ";")

# Check structure of data
str(wine)

# Distribution plots to check skewness and outliers
ggplot(wine, aes(x = alcohol)) + geom_histogram(bins = 30, fill = "skyblue") + ggtitle("Distribution of Alcohol")
ggplot(wine, aes(x = volatile.acidity)) + geom_histogram(bins = 30, fill = "lightcoral") + ggtitle("Distribution of Volatile Acidity")
ggplot(wine, aes(x = pH)) + geom_histogram(bins = 30, fill = "lightgreen") + ggtitle("Distribution of pH")

# Check for missing values
cat("Missing values:", sum(is.na(wine)))
```

## What type of data do these attributes hold?
Data Type	Attributes
**Nominal** (Categorical, Unordered): no variables
**Ordinal** (Categorical, Ordered): quality, quality_factor
**Interval**: pH
**Ratio**: fixed acidity, volatile acidity, citric acid, residual sugar, 
chlorides, free sulfur dioxide, total sulfur dioxide, density, sulphates, 
and alcohol

## -- Data Preparation --
## How was the data loaded? Were any changes made to it?
The data was loaded from a CSV file (`winequality-red.csv`). No missing values 
were found. It created a categorical version of `quality` for boxplot visualization, 
and created a binary label `good` = 1 if quality >= 7, else 0

```{r}
# Basic dimensions
cat("Number of samples: ", nrow(wine), "\n")
cat("Number of features (excluding quality): ", ncol(wine)-1, "\n")

# View column names
colnames(wine)

# Summary statistics for all features
summary(wine)
head(wine)

# Data types of features
sapply(wine, class)

# Manual classification of feature types
feature_types <- tibble(Feature = names(wine), 
                        Role = ifelse(names(wine) == "quality", "Target", "Predictor"), 
                        Type = ifelse(names(wine) == "quality", "Ordinal", "Ratio"))
print(feature_types)

# Create factor version of quality for plotting and testing
wine$quality_factor <- factor(wine$quality)
```

## -- EDA: Exploratory Data Analysis & Modeling -- 
## Correlation Matrix and Scatterplot Matrix
- **Alcohol:** Positively correlated with quality
- **Volatile Acidity:** Negatively correlated with quality
- **Density:** Negatively correlated with alcohol

We want to understand how each chemical property relates to wine quality. A correlation 
matrix shows the strength and direction of relationships between variables. 

```{r}
# Correlation matrix
corrplot(cor(wine[, 1:11]), method = "color", tl.cex = 0.8)
# Scatterplot matrix
GGally::ggpairs(wine[, c("alcohol", "volatile.acidity", "sulphates", "citric.acid", "quality")])

```
From this correlation matrix, we observed that alcohol has a positive correlation 
with quality, while volatile acidity has a negative correlation. 

## Boxplots by Wine Quality
```{r boxplots}
p1 <- ggplot(wine, aes(x = quality_factor, y = alcohol)) + geom_boxplot() + ggtitle("Alcohol vs Quality")
p2 <- ggplot(wine, aes(x = quality_factor, y = volatile.acidity)) + geom_boxplot() + ggtitle("Volatile Acidity vs Quality")

grid.arrange(p1, p2, ncol = 2)
```

## Histogram to show distribution of key variables
```{r histograms}
ggplot(wine, aes(x = alcohol)) + geom_histogram(bins = 30) + ggtitle("Distribution of Alcohol")
```

## Hypothesis Testing
## What is my hypothesis?
H0: Mean alcohol content of quality 5 and quality 7 wines are equal
Ha: Mean alcohol content of quality 7 wines is higher

T-test shows statistically significant difference in **alcohol content** between 
wines rated 5 vs. 7 (`p < 0.05`)

```{r t-test}
# Check normality assumptions for each group
shapiro.test(wine$alcohol[wine$quality == 5])
shapiro.test(wine$alcohol[wine$quality == 7])

# Check variance equality
var.test(wine$alcohol[wine$quality == 5], wine$alcohol[wine$quality == 7])

# Running t-test
t.test(alcohol ~ quality_factor, data = wine %>% filter(quality %in% c(5, 7)))
```

We checked the normality assumption for each group using the Shapiro-Wilk test. 
Both groups show slight deviation from normality (p < .05), but the t-test is 
strong to normality violations with large samples. We also check for equal variance 
using the F-test. 

## Logistic Regression (Good vs. Bad Wine)
Predicts whether a wine is "good" (`quality` >= 7) based on the selected chemical properties. 
Using alcohol, volatile acidity, sulphates, and citric acid as predictors. 
```{r logistic-model}
wine$good <- ifelse(wine$quality >= 7, 1, 0)

model <- glm(good ~ alcohol + volatile.acidity + sulphates + citric.acid, data = wine, family = binomial)
summary(model)
```

## -- Actionable Insights --
## What are my observations?
Running a correlational analysis, my observations are that there is a strong 
negative correlation between alcohol and density, and alcohol has a strong 
positive correlation with quality. 
1. **Increase Alcohol Content**
  Strongest predictor of higher quality
2. **Lower Volatile Acidity**
  Associated with better taste and aroma
3. **Optimize Sulphates and Citric Acid**
  Can improve preservation and perceived freshness
